On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2014-03-13 tex/PaperI/ms.tex
some Paper II updates + accepted version of Paper I

In the core accretion hypothesis, giant planets form by gas accretion onto solid protoplanetary cores.  The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. The actual value depends on several factors: the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids. Motivated by ongoing direct imaging searches for giant planets, this study investigates core mass requirements in the outer disk.  To determine the fastest allowed rates of gas accretion, we consider solid cores that no longer accrete planetesimals, as this would heat the gaseous envelope. Our spherical, two-layer atmospheric cooling model includes an inner convective region and an outer radiative zone that matches onto the disk.  We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard \textbf{interstellar grain} opacities.  Lower temperatures in the outer disk explain this trend, while variations in disk density are less influential.  At all distances, a lower dust opacity or higher mean molecular weight reduces the critical core mass. Our non-self-gravitating, analytic cooling model reveals that self-gravity significantly affects early atmospheric evolution, starting when the atmosphere is only $\sim$$10\%$ as massive as the core.

@@

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric, remains in hydrostatic balance, and matches onto the disk's midplane temperature and pressure at the planet's Hill radius.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
%\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions in the disk midplane.
\item Gravitational contraction of the atmosphere is the only source of planetary luminosity.  
\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. \textbf{We justify this assumption in \S\ref{sec:twolayer} and confirm its validity in \S\ref{sec:endoftime}.} 
%\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item The atmosphere obeys a polytropic equation of state (EOS), \textbf{with adiabatic index $\gamma = 7/5$ for an ideal diatomic gas.  Corrections from a realistic EOS are discussed in \S\ref{sec:EOS} and deferred to future work.}
\item Dust grains provide the opacity in the radiative zone. \textbf{For the cool temperatures in the outer disk, the radiative layer remains cool enough to avoid dust sublimation.} Other opacity choices are discussed in \S\ref{sec:op}.
%\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
\end{enumerate}
%The remainder of this section develops our model in more detail.

@@

We quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.  \textbf{Thus when $\RB < \RH$, our computational domain -- which always ends at $\RH$ -- extends further than the location where we define atmosphere mass, within $\RB$.  We emphasize that this mass definition does not affect the evolutionary calculation, which consistently includes the enclosed mass at all radii.   We justify our mass convention as follows.  First, we want to conservatively define planet mass in a way that does not exaggerate atmospheric growth and is more consistent with previous works that use $\RB$ as the outer boundary in this regime (e.g., \citealt{ikoma00}, \citealt{pollack96}).  Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.}

\textbf{The choice of $\RH$ as the outer boundary is based on this being the largest radius where spherical hydrostatic balance is approximately (but not exactly, see \S\ref{sec:hydro}) valid.  When $\RB < \RH$, we thus include the fact that the density at $\RB$ slightly exceeds the background disk density (see R06).  In practice, this effect is not very significant.  We would get similar results by choosing the outer boundary at $\RB$ in this regime, as previous studies have done.}

@@

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone, \textbf{i.e. the RHS of \Eq{eq:structd} is set to zero.  This assumption greatly simplifies the numerical problem.  Instead of solving time dependent partial differential equations, we can solve for  static solutions to a set of ODEs, Equations (\ref{eq:struct}a -- c), which we connect in a time series as described below. We show in \S\ref{sec:endoftime} that the approximation of constant luminosity in the outer layer holds for our regime of interest.  }

@@

To obtain a single atmosphere solution (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  A luminosity value is required to compute $\delrad$ and integrate Equations (\ref{eq:struct}a--c).  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$. \textbf{Physically, the luminosity determines the location of the radiative-convective boundary (RCB), consistent with the structure equations and the Schwarzschild criterion.}

\textbf{To understand time evolution, we construct} an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.  We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.  To establish the time difference between neighboring solutions, we apply \Eq{eq:coolingglobal} at the RCB.  In principle, energy balance could be evaluated at any level.  Our approximate treatment of the radiative zone makes the RCB  the preferred location.     The elapsed time $\Delta t$  between states $i$ and $i +1$ is given by the finite difference
\begin{equation}
\label{eq:dti}
\Delta t = %\left(
{ -\Delta E + \brak{e} \Delta M - \brak{P}  \Delta V_{\brak{M}} \over \brak{L} }\, ,
\end{equation} 
using \Eq{eq:coolingglobal} with $\Gamma = L\co = 0$.
Brackets indicate an average of, and $\Delta$ indicates a difference between, the two states.  All values are evaluated at the RCB.  Due to the partial derivative in \Eq{eq:coolingglobal}, the volume difference $\Delta V_{<M>}$ is performed at fixed mass, here the average of the masses at the RCB.

@@

The analytic model also assumes that the upper radiative layer is thick enough that $P\cb \gg P\di$.  This approximation ignores the earliest stages of cooling, which are rapid enough to be of minor importance.  The analytic model also ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.  \textbf{Finally, while the numerical models set the outer boundary at $\RH$, we simplify the analytic calculations by setting the outer boundary to infinity, effectively neglecting the finite scale-height of the disk.   As shown in \App{iso}, the effect of this approximation is minor for $\RB \lesssim \RH$, the primary case of interest (see also R06).}

\textbf{Of all the approximations, the neglect of self-gravity is by far the most significant, as we have verified by comparison to numerical integrations that only neglect self-gravity.}

@@

To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.   Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway growth criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$, with $f < 1$.  Specifically, we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.   This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.  Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.  Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.}  With $f = 0.13$, the revised scalings are
\begin{subeqnarray}
t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
\end{subeqnarray} 
now using $t_{\rm run}$, defined as the time when $M_{\rm atm} = f M\co$,  instead of $t_{\rm co}$ for the runaway growth timescale.  When plotting these rescaled analytic results, we include the variations in $\xi$ from \Eq{eq:xidef}.

@@

\textbf{The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, thermal energy is a fixed fraction of the gravitational potential energy, giving $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$  as in \Eq{eq:deep}.  This behavior is seen in \Fig{fig:profiles} for $r\ll R\cb$. Near the core, \Eq{eq:Tconvdeep} gives the core temperature, $T\co= G M\co / (C_{\rm P} R\co)$, unaffected by  the overlying atmosphere mass which does not contribute to the gravitational potential.  Closer to the RCB, self-gravity is no longer negligible, particularly for large envelope masses. Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.}

@@

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 8.99 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}. \textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with atmospheric mass (cf. Equations \ref{eq:PcbRcb} and \ref{eq:crit}).}

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.  \textbf{Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.  Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.  Nevertheless, all our models show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}

@@

%We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
\textbf{We now use our structure and evolution models to estimate the timescales and minimum core masses for giant planet formation for a range of disk conditions and other model parameters.  Our results for atmosphere growth times -- the time for a core of fixed mass to undergo runaway gas accretion -- are presented in \S\ref{sec:tcross}.  Section \ref{sec:critcore} gives our results for critical core masses, the minimum values that trigger runaway atmospheric growth within a plausible disk lifetime, here 3 Myr.}

@@

The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.  A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.  \textbf{This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.   Moreover, the Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).  To see how this affects cooling times, note that the characteristic RCB depth near runaway scales as $P\cb \sim P_{\rm M} \propto \mu^{-4}$ from \Eq{eq:PM}.  Thus $L \propto 1/P\cb \propto \mu^4$ explains the trend of slower cooling for lower $\mu$.}
%he larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf. \Eq{eq:PcbRcb}.}

For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.  Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales. \textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here (see \S\ref{sec:EOS}).}

@@

\textbf{A larger disk lifetime further reduces $M_{\rm crit}$. Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}. As $M_{\rm crit} \sim t\di^{-3/5}$ (cf. Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.  Of course, some fraction of the disk lifetime must be allocated to the growth and possibly migration of the core.}

@@

\subsection{\textbf{Comparison with Previous Studies}}
\label{sec:comp}

\textbf{We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.  A major distinguishing feature of our study is that we explore a range of disk conditions, while I00 and PN05 focus on growth at the location of Jupiter, 5.2 AU, in their disk models.  Moreover, both I00 and PN05 solve the full time dependent energy equation and consider more detailed EOSs, in addition to other detailed differences in model parameters.  Despite these differences, the qualitative and quantitative agreement with our study is good.}

\textbf{I00 give an approximate analytic fit to their models' envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), at $a=5.2$ AU, }
\begin{equation}
\label{eq:tauenv}
\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr}\, .
\end{equation}
\textbf{Our model reproduces the linear opacity dependence (see \S\ref{sec:timeev}), even though we use a temperature dependent opacity, \Eq{eq:opacitylaw}, instead of a constant $\kappa$.  Our growth times at 5 AU,} 
\begin{equation}
\label{eq:trun5au}
t_{\rm run} \sim 5 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.4} \,\,\, \text{yr},
\end{equation} 
\textbf{agree well with the I00 results, both in magnitude and scaling with core mass.}

\textbf{One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models, as described in the introduction.  Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) shows that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.  We not only reproduce this effect in our simplified model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}), but we also show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.}

\textbf{Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.  This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool. \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity. These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase. The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.}

\textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.  Note, however, that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}

\textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.  These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.}

\section{Neglected Effects}\label{sec:neglected}
\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored   effects of potential significance.  We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend our current models to address some of these effects.}

>>>>

In the core accretion hypothesis, giant planets form by gas accretion onto solid protoplanetary cores.  The minimum (or critical) core mass to form a gas giant is typically quoted as $10 M_{\oplus}$. The actual value depends on several factors: the location in the protoplanetary disk, atmospheric opacity, and the accretion rate of solids. Motivated by ongoing direct imaging searches for giant planets, this study investigates core mass requirements in the outer disk.  To determine the fastest allowed rates of gas accretion, we consider solid cores that no longer accrete planetesimals, as this would heat the gaseous envelope. Our spherical, two-layer atmospheric cooling model includes an inner convective region and an outer radiative zone that matches onto the disk.  We determine the minimum core mass for a giant planet to form within a typical disk lifetime of 3 Myr.   The minimum core mass declines with disk radius, from $\sim$$8.5 M_{\oplus}$ at 5 AU to $\sim$$3.5 M_{\oplus}$ at 100 AU, with standard interstellar grain opacities.  Lower temperatures in the outer disk explain this trend, while variations in disk density are less influential.  At all distances, a lower dust opacity or higher mean molecular weight reduces the critical core mass. Our non-self-gravitating, analytic cooling model reveals that self-gravity significantly affects early atmospheric evolution, starting when the atmosphere is only $\sim$$10\%$ as massive as the core.

@@

Our main assumptions are summarized as follows:
\begin{enumerate}
\item The atmosphere is spherically symmetric, remains in hydrostatic balance, and matches onto the disk's midplane temperature and pressure at the planet's Hill radius.
\item The core mass and radius are fixed in evolutionary calculations, neglecting ongoing planetesimal or dust accretion.
%\item At the planet's Hill radius, the atmospheric temperature and pressure match the conditions in the disk midplane.
\item Gravitational contraction of the atmosphere is the only source of planetary luminosity.  
\item Cooling of the atmosphere is dominated by the convective interior.  Thus the luminosity in the radiative exterior is held spatially constant. We justify this assumption in \S\ref{sec:twolayer} and confirm its validity in \S\ref{sec:endoftime}.
%\item A global cooling model connects independent static solutions into a time-dependent sequence.
\item The atmosphere obeys a polytropic equation of state (EOS), with adiabatic index $\gamma = 7/5$ for an ideal diatomic gas.  Corrections from a realistic EOS are discussed in \S\ref{sec:EOS} and deferred to future work.
\item Dust grains provide the opacity in the radiative zone. For the cool temperatures in the outer disk, the radiative layer remains cool enough to avoid dust sublimation. Other opacity choices are discussed in \S\ref{sec:op}.
%\item Because gas accretion accelerates after the crossover mass, the time to reach the crossover mass, i.e. the crossover time, is a good approximation of the total time to form a gas giant.
\end{enumerate}
%The remainder of this section develops our model in more detail.

@@

We quote planet masses as the enclosed mass within the smaller of $\RB$ and $\RH$.  Thus when $\RB < \RH$, our computational domain -- which always ends at $\RH$ -- extends further than the location where we define atmosphere mass, within $\RB$.  We emphasize that this mass definition does not affect the evolutionary calculation, which consistently includes the enclosed mass at all radii.   We justify our mass convention as follows.  First, we want to conservatively define planet mass in a way that does not exaggerate atmospheric growth and is more consistent with previous works that use $\RB$ as the outer boundary in this regime (e.g., \citealt{ikoma00}, \citealt{pollack96}).  Second, most of the gas outside $\RB$ is weakly bound, and unlikely to remain attached to the planet if the disk suddenly dissipates.

The choice of $\RH$ as the outer boundary is based on this being the largest radius where spherical hydrostatic balance is approximately (but not exactly, see \S\ref{sec:hydro}) valid.  When $\RB < \RH$, we thus include the fact that the density at $\RB$ slightly exceeds the background disk density (see R06).  In practice, this effect is not very significant.  We would get similar results by choosing the outer boundary at $\RB$ in this regime, as previous studies have done.

@@

In the radiative zone,  local energy balance, \Eq{eq:structd}, does affect the atmospheric structure.  We proceed by assuming that the majority of energy is lost from the convective interior, and thus the luminosity can be treated as constant in the outer radiative zone, i.e. the RHS of \Eq{eq:structd} is set to zero.  This assumption greatly simplifies the numerical problem.  Instead of solving time dependent partial differential equations, we can solve for  static solutions to a set of ODEs, Equations (\ref{eq:struct}a -- c), which we connect in a time series as described below. We show in \S\ref{sec:endoftime} that the approximation of constant luminosity in the outer layer holds for our regime of interest.

@@

To obtain a single atmosphere solution (indexed by $i$), we choose a planet mass $M_i$.  At the outer boundary, at $\RH(M_i)$, the temperature and pressure are set to the disk values.  A luminosity value is required to compute $\delrad$ and integrate Equations (\ref{eq:struct}a--c).  The correct value of the luminosity is not known in advance, and is the eigenvalue of the problem.  The boundary conditions can only be satisfied for the correct value of the luminosity eigenvalue, which we find by the shooting method.  Specifically, we alter the luminosity until the integrated value of mass at the core, $m(R\co)$, matches the actual core mass, $M\co$. Physically, the luminosity determines the location of the radiative-convective boundary (RCB), consistent with the structure equations and the Schwarzschild criterion.

To understand time evolution, we construct an array of solutions to Equations (\ref{eq:struct}a -- c) in order of increasing atmospheric mass.  We then use global energy balance, \Eq{eq:coolingglobal}, to ``follow the mass" and place these solutions in a cooling sequence.  To establish the time difference between neighboring solutions, we apply \Eq{eq:coolingglobal} at the RCB.  In principle, energy balance could be evaluated at any level.  Our approximate treatment of the radiative zone makes the RCB  the preferred location.     The elapsed time $\Delta t$  between states $i$ and $i +1$ is given by the finite difference
\begin{equation}
\label{eq:dti}
\Delta t = %\left(
{ -\Delta E + \brak{e} \Delta M - \brak{P}  \Delta V_{\brak{M}} \over \brak{L} }\, ,
\end{equation} 
using \Eq{eq:coolingglobal} with $\Gamma = L\co = 0$.
Brackets indicate an average of, and $\Delta$ indicates a difference between, the two states.  All values are evaluated at the RCB.  Due to the partial derivative in \Eq{eq:coolingglobal}, the volume difference $\Delta V_{<M>}$ is performed at fixed mass, here the average of the masses at the RCB.

@@

The analytic model also assumes that the upper radiative layer is thick enough that $P\cb \gg P\di$.  This approximation ignores the earliest stages of cooling, which are rapid enough to be of minor importance.  The analytic model also ignores the surface terms in \Eq{eq:coolingglobal}, which we show to be a modest correction in \S\ref{sec:endoftime}.  Finally, while the numerical models set the outer boundary at $\RH$, we simplify the analytic calculations by setting the outer boundary to infinity, effectively neglecting the finite scale-height of the disk.   As shown in \App{iso}, the effect of this approximation is minor for $\RB \lesssim \RH$, the primary case of interest (see also R06).

Of all the approximations, the neglect of self-gravity is by far the most significant, as we have verified by comparison to numerical integrations that only neglect self-gravity.

@@

To facilitate comparison with our numerical results, we rescale the analytic results as follows.   Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway growth criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$, with $f < 1$.  Specifically, we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.   This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.  Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.  Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.  With $f = 0.13$, the revised scalings are
\begin{subeqnarray}
t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
\end{subeqnarray} 
now using $t_{\rm run}$, defined as the time when $M_{\rm atm} = f M\co$,  instead of $t_{\rm co}$ for the runaway growth timescale.  When plotting these rescaled analytic results, we include the variations in $\xi$ from \Eq{eq:xidef}.

@@

The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, thermal energy is a fixed fraction of the gravitational potential energy, giving $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$  as in \Eq{eq:deep}.  This behavior is seen in \Fig{fig:profiles} for $r\ll R\cb$. Near the core, \Eq{eq:Tconvdeep} gives the core temperature, $T\co= G M\co / (C_{\rm P} R\co)$, unaffected by  the overlying atmosphere mass which does not contribute to the gravitational potential.  Closer to the RCB, self-gravity is no longer negligible, particularly for large envelope masses. Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.

@@

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 8.99 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}. This shift does not occur in the analytic solution, where $R\cb$ continually decreases with atmospheric mass (cf. Equations \ref{eq:PcbRcb} and \ref{eq:crit}).

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.  Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.  Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.  Nevertheless, all our models show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.

@@

%We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
We now use our structure and evolution models to estimate the timescales and minimum core masses for giant planet formation for a range of disk conditions and other model parameters.  Our results for atmosphere growth times -- the time for a core of fixed mass to undergo runaway gas accretion -- are presented in \S\ref{sec:tcross}.  Section \ref{sec:critcore} gives our results for critical core masses, the minimum values that trigger runaway atmospheric growth within a plausible disk lifetime, here 3 Myr.

@@

The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.  A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.  This effect is both well established in core accretion studies \citep{stevenson82} and intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.   Moreover, the Bondi radius decreases as $\RB \propto \mu$, giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).  To see how this affects cooling times, note that the characteristic RCB depth near runaway scales as $P\cb \sim P_{\rm M} \propto \mu^{-4}$ from \Eq{eq:PM}.  Thus $L \propto 1/P\cb \propto \mu^4$ explains the trend of slower cooling for lower $\mu$.
%he larger atmospheric scale height due to a lower $\mu$ also results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf. \Eq{eq:PcbRcb}.}

For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.  Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales. In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here (see \S\ref{sec:EOS}).

@@

A larger disk lifetime further reduces $M_{\rm crit}$. Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}. As $M_{\rm crit} \sim t\di^{-3/5}$ (cf. Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.  Of course, some fraction of the disk lifetime must be allocated to the growth and possibly migration of the core.

@@

\subsection{Comparison with Previous Studies}
\label{sec:comp}

We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.  A major distinguishing feature of our study is that we explore a range of disk conditions, while I00 and PN05 focus on growth at the location of Jupiter, 5.2 AU, in their disk models.  Moreover, both I00 and PN05 solve the full time dependent energy equation and consider more detailed EOSs, in addition to other detailed differences in model parameters.  Despite these differences, the qualitative and quantitative agreement with our study is good.

I00 give an approximate analytic fit to their models' envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), at $a=5.2$ AU, 
\begin{equation}
\label{eq:tauenv}
\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr}\, .
\end{equation}
Our model reproduces the linear opacity dependence (see \S\ref{sec:timeev}), even though we use a temperature dependent opacity, \Eq{eq:opacitylaw}, instead of a constant $\kappa$.  Our growth times at 5 AU,
\begin{equation}
\label{eq:trun5au}
t_{\rm run} \sim 5 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.4} \,\,\, \text{yr},
\end{equation} 
agree well with the I00 results, both in magnitude and scaling with core mass.

One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models, as described in the introduction.  Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) shows that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.  We not only reproduce this effect in our simplified model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}), but we also show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.

Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.  This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool. \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity. These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase. The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.

A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.  Note, however, that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.

In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.  These studies consider the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance claimed by \cite{Raf11} (and references therein) can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  Nevertheless, the plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains uncertain, as discussed in more detail in \S\ref{sec:conclusions}.

\section{Neglected Effects}\label{sec:neglected}
Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored   effects of potential significance.  We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend our current models to address some of these effects.
